Davos: a Python package ``smuggler'' for constructing lightweight reproducible notebooks

### 2023-09-25 paper/main.tex
add ref for scikit-learn docs page
fixed a typo in section 2.2.5
update fig 3 caption, other minor wording changes & typo fixes pre-section 3

The Davos package also provides a handful of functions available in the top-level \texttt{davos} namespace.
Some of these function serve primarily as conveniences, while others provide additional functionality:

@@

\item \texttt{require\_python(version\_spec, warn=False, extra\_msg=None, prereleases=None)}: Through \texttt{smuggle} statements and onion comments, Davos can automatically ensure that all Python packages needed to run a notebook are installed, and that the same versions of those packages are used no matter when or by whom the notebook is run.
  However, because Davos operates at runtime, one thing it cannot do automatically is install and switch to a specific version of Python itself.
  Distributing shared code along with a precise Python version for running it requires a heavier-weight solution, such as a Conda environment or Docker container (see Fig.~\ref{fig:code-sharing}).
  Yet a Davos-enhanced notebook may still \texttt{smuggle} certain packages that depend on users having a particular Python version or range of versions (e.g., even just within the standard library, the \texttt{dataclass} module was first added in Python 3.7 \cite{Smit17} and at least 19 modules are slated for removal in Python 3.13 \cite{HeimCann19}).
  The \texttt{davos.require\_python()} function can be added to the top of a Davos-enhanced notebook to communicate to users that the notebook's code should be run with a specific or constrained Python version (see Sec.~\ref{sec:illustrative-example} for example usage).
  The function may be passed a version identifier (e.g., \texttt{"3.10.5"}) or any valid version specifier \cite{CoghStuf13} (e.g., \texttt{"\raisebox{0.5ex}{\texttildelow}=3.11"}, \texttt{">=3.9;<3.12"}, etc.) and will raise an exception if the user's Python version is incompatible.
  Alternatively, a ``soft'' or suggested constraint can be specified by passing \texttt{warn=True} to issue a warning rather than raise an error.
  Additional information can be added to the default error/warning message (e.g., the specific reason for this requirement) via the \texttt{extra\_msg} argument, and the optional \texttt{prereleases} argument can be used to explicitly allow (\texttt{True}) or disallow (\texttt{False}) pre-release versions (by default, the policy is determined by the value of \texttt{version\_spec}).

@@

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/flow_chart}
\caption{\small \textbf{\texttt{smuggle()} function algorithm.}  At a
  high level, the \texttt{smuggle()} function may be conceptualized as
following two basic steps.  First (left), Davos ensures that the
correct version of the desired package has been installed, carrying
out the installation automatically if needed.  Second (right),
Davos imports the package and updates the current runtime environment.}
\label{fig:flow-chart}
\end{figure}

@@

The example code throughout Section \ref{subsec:onion} illustrates how Davos is most typically used:
By including a series of \texttt{smuggle} statements and onion comments with version specifiers or other options in an IPython notebook, researchers can share their code and its dependencies in a single file that can be easily run without any additional setup, creates and manages its own isolated environment, automatically installs its required packages at runtime, and ensures that the package versions with which it is run do not change unexpectedly.
In this section, we have contrived a more complex scenario to highlight some of Davos's more advanced features and illustrate how they may be used to handle certain challenges that can arise with sharing reproducible scientific code.

\bigskip\bigskip\textcolor{red}{\textbf{========== TODO: finish editing from here to end ==========}}\bigskip\bigskip

Across different versions of a given package, various modules, functions,
and other objects may be updated, removed, renamed, or otherwise altered. In
addition to changing the behaviors of active computations, these changes can
render saved objects created using one version of a package incompatible with
other versions of the same package. For example, the popular
\texttt{pandas}~\cite{McKi10} library used to include the \texttt{Panel} data
structure for storing 3-dimensional arrays. Since version 0.20.0, however, the
\texttt{Panel} class has been deprecated, and in version 0.25.0, it was removed
entirely. Suppose a user had a dataset stored in a \texttt{Panel} object
(created using an older version of \texttt{pandas}) and had saved it to their
disk (e.g., for later reuse or to share with other users) by serializing the
\texttt{Panel} with Python's \texttt{pickle} protocol. The \texttt{pickle}
protocol is a popular built-in method of persisting data in Python that allows
users to save, share, and load arbitrary objects. However, in order to
successfully ``unpickle'' (i.e., load and restore) a ``pickled'' (i.e., previously saved)
object, the object's class must be defined in and importable from the same
module as when it was saved. Thus, because of the \texttt{Panel} class's
removal, the user's dataset could not be read by any version of \texttt{pandas}
from 0.25.0 or beyond. These incompatibilities are also not limited solely to
traditional forms of data. For example, saved model states and other objects
may reference modules, functions, attributes, classes, or other objects that
may not be identical (or even present) across all versions of their associated
package.

The example provided in Figure~\ref{fig:illustrative-example} demonstrates how
the Davos package can be used to circumvent these incompatibilities by
carefully controlling which versions of each package are used in different
parts of the notebook. The example shows how a dataset and model that require
now-incompatible components of the \texttt{pandas} and
\texttt{scikit-learn}~\cite{PedrEtal11} packages may be loaded in (using older
versions of each package) and used alongside more recent versions of each
package that provide new and improved functionality. When included at the top
of a Jupyter notebook, the code in Figure~\ref{fig:illustrative-example}
ensures that these objects will be loaded successfully and analyzed using the
same set of package versions, no matter when or by whom the notebook is run.

@@

Next, we load in the reformatted dataset (line 33) and pre-trained
model (line 35) that we wish to use in our analysis.  In our
hypothetical example, we can suppose that the model was provided as a
\texttt{scikit-learn} \texttt{Pipeline} object that passes data
through two pretrained models in succession.  First, a trained \texttt{CountVectorizer}
instance converts text data to an array of word counts.  Second, the
word counts are passed to a topic model~\cite{BleiEtal03} using a
pretrained \texttt{LatentDirichletAllocation} instance.
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example7}
\end{center}
Let us suppose that the \texttt{Pipeline} object had been saved by its
original creator using the \texttt{joblib} package, as
\texttt{scikit-learn}'s documentation recommends.  Because
\texttt{joblib} uses the \texttt{pickle} protocol internally, the
ability to save and load pre-trained models is not guaranteed across
different \texttt{scikit-learn} versions.  For example, suppose that
the \texttt{Pipeline} object was created using \texttt{scikit-learn}
v0.21.3.  In that version of \texttt{scikit-learn}, the
\texttt{LatentDirichletAllocation} class was defined in
\texttt{sklearn.decomposition.online\_lda}.  However, in version
0.22.0, that module was renamed to \texttt{\_online\_lda}, and in
version 0.22.1, it was again renamed to \texttt{\_lda}.

>>>>

The Davos package also provides a handful of functions available in the top-level \texttt{davos} namespace.
Some of these functions serve primarily as conveniences, while others provide additional functionality:

@@

\item \texttt{require\_python(version\_spec, warn=False, extra\_msg=None, prereleases=None)}: Through \texttt{smuggle} statements and onion comments, Davos can automatically ensure that all Python packages needed to run a notebook are installed, and that the same versions of those packages are used no matter when or by whom the notebook is run.
  However, because Davos operates at runtime, one thing it cannot do automatically is install and switch to a specific version of Python itself.
  Distributing shared code along with a precise Python version for running it requires a heavier-weight solution, such as a Conda environment or Docker container (see Fig.~\ref{fig:code-sharing}).
  Yet a Davos-enhanced notebook may still \texttt{smuggle} certain packages that depend on users having a particular Python version or range of versions (e.g., even just within the standard library, the \texttt{dataclass} module was first added in Python 3.7 \cite{Smit17} and at least 19 modules are slated for removal in Python 3.13 \cite{HeimCann19}).
  The \texttt{davos.require\_python()} function can be added to the top of a Davos-enhanced notebook to communicate to users that the notebook's code should be run with a specific or constrained Python version (see Sec.~\ref{sec:illustrative-example} for example usage).
  The function may be passed a version identifier (e.g., \texttt{"3.10.5"}) or any valid version specifier \cite{CoghStuf13} (e.g., \texttt{"\raisebox{0.5ex}{\texttildelow}=3.11"}, \texttt{">=3.9;<3.12"}, etc.) and will raise an exception if the user's Python version is incompatible.
  Alternatively, a ``soft'' or suggested constraint can be imposed by passing \texttt{warn=True} to issue a warning rather than raise an error.
  Additional information can be added to the default error/warning message (e.g., the specific reason for this requirement) via the \texttt{extra\_msg} argument, and the optional \texttt{prereleases} argument can be used to explicitly allow (\texttt{True}) or disallow (\texttt{False}) pre-release versions (by default, the policy is determined by the value of \texttt{version\_spec}).

@@

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/flow_chart}
\caption{\small \textbf{\texttt{smuggle()} function algorithm.}  At a
  high level, the \texttt{smuggle()} function may be conceptualized as
  following two basic steps.  First (left), Davos ensures that the
  correct version of the desired package is available locally, installing
  it automatically (into the notebook's project directory) if needed.  Second (right),
  Davos loads the package into the notebook and updates the current
  runtime environment.}
\label{fig:flow-chart}
\end{figure}

@@

The example code throughout Section \ref{subsec:onion} illustrates how Davos is most typically used:
By including a series of \texttt{smuggle} statements and onion comments with version specifiers or other options in an IPython notebook, researchers can share their code and its dependencies in a single file that can be easily run without any additional tools or setup, creates and manages its own isolated environment, automatically installs its required packages at runtime, and ensures that the package versions with which it is run do not change unexpectedly.
In this section, we have contrived a more complex scenario to highlight some of Davos's more advanced features, and illustrate how they may be used to handle certain challenges that can arise when writing and sharing reproducible scientific code.

Across different versions of a given package, various modules, functions,
and other objects may be updated, removed, renamed, or otherwise altered. In
addition to changing the behaviors of active computations, these changes can
render saved objects created using one version of a package incompatible with
other versions of the same package. For example, the popular
\texttt{pandas}~\cite{McKi10} library used to include the \texttt{Panel} data
structure for storing 3-dimensional arrays. Since version 0.20.0, however, the
\texttt{Panel} class has been deprecated, and in version 0.25.0, it was removed
entirely. Suppose a user had a dataset stored in a \texttt{Panel} object
(created using an older version of \texttt{pandas}) and had saved it to their
disk (e.g., for later reuse or to share with other users) by serializing the
\texttt{Panel} with Python's \texttt{pickle} protocol. The \texttt{pickle}
protocol is a popular built-in method of persisting data in Python that allows
users to save, share, and load arbitrary objects. However, in order to
successfully ``unpickle'' (i.e., load and restore) a ``pickled'' (i.e., previously saved)
object, that object's class must be defined in and importable from the same
module as it was when the object was originally saved. Thus, because of the \texttt{Panel} class's
removal, the user's dataset could not be read by any version of \texttt{pandas}
from 0.25.0 onward. These incompatibilities are also not limited solely to
traditional forms of data. For example, saved model states and other objects
may reference modules, functions, attributes, classes, or other objects that
may not be identical (or even present) across all versions of their associated
packages.

The example provided in Figure~\ref{fig:illustrative-example} demonstrates how
Davos can be used to circumvent these incompatibilities by
temporarily switching between different versions of the same package within a single runtime.
%carefully controlling which versions of each package are used in different parts of the notebook.
The example shows how a dataset and model that require
now-incompatible components of the \texttt{pandas} and
\texttt{scikit-learn}~\cite{PedrEtal11} libraries may be loaded in (using older
versions of each package) and used alongside more recent versions of each
package that provide new and improved functionality. When included at the top
of an IPython notebook, the code in Figure~\ref{fig:illustrative-example}
ensures that these objects will be loaded successfully and analyzed using the
same set of package versions, no matter when or by whom the notebook is run.

@@

Next, we load in the reformatted dataset (line 33) and pre-trained
model (line 35) that we wish to use in our analysis.  In our
hypothetical example, we can suppose that the model was provided as a
\texttt{scikit-learn} \texttt{Pipeline} object that passes data
through two pretrained models in succession.  First, a trained \texttt{CountVectorizer}
instance converts text data to an array of word counts.  Second, the
word counts are passed to a topic model~\cite{BleiEtal03} using a
pretrained \texttt{LatentDirichletAllocation} instance.
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example7}
\end{center}
Let us suppose that the \texttt{Pipeline} object had been saved by its
original creator using the \texttt{joblib} package, as
\texttt{scikit-learn}'s documentation recommends \cite{skle22}.  Because
\texttt{joblib} uses the \texttt{pickle} protocol internally, the
ability to save and load pre-trained models is not guaranteed across
different \texttt{scikit-learn} versions.  For example, suppose that
the \texttt{Pipeline} object was created using \texttt{scikit-learn}
v0.21.3.  In that version of \texttt{scikit-learn}, the
\texttt{LatentDirichletAllocation} class was defined in
\texttt{sklearn.decomposition.online\_lda}.  However, in version
0.22.0, that module was renamed to \texttt{\_online\_lda}, and in
version 0.22.1, it was again renamed to \texttt{\_lda}.
